Supplement to Charity During Crackdown: Analyzing the Impact of State Repression of NGOs on Philanthropy

### 2020-01-21 manuscript/manuscript.Rmd
More final-ish edits

---
title: "Charity During Crackdown: Analyzing the Impact of State Repression of NGOs on Philanthropy"
short-title: Charity During Crackdown
author:
- name: Suparna Chaudhry
  affiliation: Christopher Newport University
  email: suparna.chaudhry@cnu.edu
  url: http://www.suparnachaudhry.com/
  twitter: SuparnaChaudhry
- name: Andrew Heiss
  affiliation: Brigham Young University
  email: andrew_heiss@byu.edu
  url: https://www.andrewheiss.com/
  twitter: andrewheiss
date: "`r format(Sys.time(), '%B %e, %Y')`"
git-repo: https://github.com/andrewheiss/ngo-crackdowns-philanthropy
reference-section-title: References
toc: false
title-page: true
abstract: >-
  State crackdown on non-governmental organizations (NGOs), particularly through legal restrictions on funding and advocacy, has become increasingly pervasive across the globe. How do legal restrictions on NGOs impact patterns of private philanthropic giving and individual donor decision-making? With reduced funding for NGOs working on contentious issues, and an absence of philanthropic culture in various developing countries, many NGOs rely on funds from private donors in Western countries. Existing research has predominantly focused on individual-level determinants of philanthropy. However, we argue that structural factors such as domestic political environments of NGO host countries can also influence preferences of foreign private donors, and that crackdown on NGOs serve as a heuristic to donors of organizational deservingness. Using a survey experiment, we explore how this crackdown changes donors' preferences based on the issue area and funding sources of the NGO. We find that while crackdowns do not directly influence the likelihood of donation on their own, donors are willing to donate substantially more to legally besieged NGOs abroad. We conclude by discussing the implications of our results on the sustainability of civil society organizations working in repressive countries. 
thanks: >-
  We thank Mark Buntaine, Michael DeCrescenzo, Jeffrey Friedman, Thomas Leeper, and Tristan Mahr for their helpful comments and input. We also thank Samantha Camilletti and Joshua Dutro for their excellent research assistance. This experiment received approval by the human subjects research committees at both Christopher Newport University (018.042) and Brigham Young University (E18104). We preregistered our hypotheses and research design at the Open Science Framework, and our preregistration protocol is available at <https://osf.io/dx973/>.
author-note: >-
  Suparna Chaudhry, PhD, Assistant Professor of Political Science, Christopher Newport University; Andrew Heiss, PhD, Assistant Professor of Public Management, Andrew Young School of Policy Studies, Georgia State University. 
additional-info: Figures should appear in color online only, not in print.
correspondence: >-
  Correspondence concerning this article should be addressed to Suparna Chaudhry, PhD, Department of Political Science, Christopher Newport University, Newport News, VA 23606, 757-594-7821. Contact: suparna.chaudhry@cnu.edu
word-count: "4,081"
keywords: international NGOs, donor behavior, philanthropy, human rights, repression
---

@@

![Causal diagram showing causes of donor behavior. Y = likelihood of donation and amount donated; C = anti-NGO legal crackdown, I = NGO issue area, F = NGO funding source.](output/figures/causal-path.pdf){#fig:causal-path}

@@

Issue area contentiousness directly influences legal crackdowns, however, and confounds the heuristic effect of crackdowns on donor behavior (see @fig:causal-path). NGOs working on more contentious human rights issues can be challenging to regimes [@Heiss:2016; @Swiney:2019]. These organizations are often viewed as politically motivated outsiders and run a higher risk of getting expelled [@DPR:2015], increasing the possibility of wasting donor resources. Thus, donors may believe that it is more difficult to improve human rights in the target country and subsequently shy away from donating to NGOs working on political causes.

@@

The final heuristic that can influence individual donors that we examine is the *source of NGO funding*. Government grants to NGOs can often crowd out private funding for nonprofit organizations [@AndreoniPayne:2003; @Brooks:2000], though evidence is mixed on whether this completely or partially crowds out private dollars [@Steinberg:1991; @Heutel:2014]. Donors may feel less inclined to contribute to NGOs receiving home government funds because such funding can signal a lack of independence from the government. Organizations that receive substantial government funding frequently avoid programming that would question the donor government even if it were in line with their mission [@Stroup:2012]. In contrast, individual donors who see that organizations are privately funded may feel that they can also contribute and help.

@@

Importantly, when testing these hypotheses, we do not assume that individual donors have perfect knowledge about an NGO or the countries an NGO works in. Donors do not always know about anti-NGO laws in target countries or who funds an NGO when they make their contributions. Moreover, the average donor will not know exactly what "legal crackdown" entails for any given organization (i.e. does it mean an NGO was expelled, or that its assets were frozen, or that it received a fine?". For the sake of this experiment, this ambiguous definition is not an issue. The heuristics we explore here are explicitly simplified shortcuts to donor information-gathering. Knowing that an organization works with human rights issues (even if they don't know the exact subfield of human rights) should be enough to trigger the issue-based deservingess heuristic. Similarly, knowing that an NGO faces legal trouble should be enough to trigger the structural deservingness heuristic.

@@

We collected two outcome measures: (1) how likely participants would be to donate to IRC (measured with a 5-point scale ranging from "Extremely likely" to "Extremely unlikely"), and (2) how much participants would hypothetically donate to IRC if they had an extra $100. We collapsed the likelihood scale into a binary variable measuring whether the participant is likely ("Extremely likely" and "Somewhat likely") or not likely (all other responses) to donate. We also included an exploratory free response question asking respondents to justify their choice.

@@

```{r child = "output/tables/tbl-likely-diffs.md"}

```

@@

```{r child = "output/tables/tbl-amount-diffs.md"}

```

@@

This research note should be viewed as an exploratory study, and it raises numerous questions for future research. These include determining the extent to which these preferences translate into behavior—here, we have only asked respondents to make a hypothetical donation, but preferences may not reflect actual behavior. Future research can also test the generalizability of our findings to other issue areas. TODO OTHER THINGS: Crackdown in authoritarian regimes, but there's an increase in restrictive laws in democracies [@Swiney:2019]

@@

[^supp]: See the supplement for details about our sample demographics and treatment assignment balance.

[^priors]: See the supplement for details about our prior choices and computational environment.

[^mturk]: TODO: Justify MTurk here.

[^targeting]: While anti-NGO legislation (in the form of barriers to entry, funding, and advocacy) can theoretically target all NGOs (or a large subset of them) within a state's borders, in practices, these laws are enforced more selectively against contentious groups that pose threats to states [see @Chaudhry:2016; @Heiss:2017].

>>>>

---
title: "Charity During Crackdown: Analyzing the Impact of State Repression of NGOs on Philanthropy"
short-title: Charity During Crackdown
author:
- name: Suparna Chaudhry
  affiliation: Christopher Newport University
  email: suparna.chaudhry@cnu.edu
  url: http://www.suparnachaudhry.com/
  twitter: SuparnaChaudhry
- name: Andrew Heiss
  affiliation: Georgia State University
  email: aheiss@gsu.edu
  url: https://www.andrewheiss.com/
  twitter: andrewheiss
date: "`r format(Sys.time(), '%B %e, %Y')`"
git-repo: https://github.com/andrewheiss/ngo-crackdowns-philanthropy
reference-section-title: References
toc: false
title-page: true
abstract: >-
  State crackdown on non-governmental organizations (NGOs), particularly through legal restrictions on funding and advocacy, has become increasingly pervasive across the globe. How do legal restrictions on NGOs impact patterns of private philanthropic giving and individual donor decision-making? With reduced funding for NGOs working on contentious issues, and an absence of philanthropic culture in various developing countries, many NGOs rely on funds from private donors in Western countries. Existing research has predominantly focused on individual-level determinants of philanthropy. However, we argue that structural factors such as domestic political environments of NGO host countries can also influence preferences of foreign private donors, and that crackdown on NGOs serve as a heuristic to donors of organizational deservingness. Using a survey experiment, we explore how this crackdown changes donors' preferences based on the issue area and funding sources of the NGO. We find that while crackdowns do not directly influence the likelihood of donation on their own, donors are willing to donate substantially more to legally besieged NGOs abroad. We conclude by discussing the implications of our results on the sustainability of civil society organizations working in repressive countries. 
thanks: >-
  We thank Mark Buntaine, Michael DeCrescenzo, Jeffrey Friedman, Thomas Leeper, and Tristan Mahr for their helpful comments and input. We also thank Samantha Camilletti and Joshua Dutro for their excellent research assistance. This experiment received approval by the human subjects research committees at both Christopher Newport University (018.042) and Brigham Young University (E18104). We preregistered our hypotheses and research design at the Open Science Framework, and our preregistration protocol is available at <https://osf.io/dx973/>.
author-note: >-
  Suparna Chaudhry, PhD, Assistant Professor of Political Science, Christopher Newport University; Andrew Heiss, PhD, Assistant Professor of Public Management, Andrew Young School of Policy Studies, Georgia State University. 
additional-info: Figures should appear in color online only, not in print.
correspondence: >-
  Correspondence concerning this article should be addressed to Suparna Chaudhry, PhD, Department of Political Science, Christopher Newport University, Newport News, VA 23606, 757-594-7821. Contact: suparna.chaudhry@cnu.edu
word-count: "4,081"
keywords: international NGOs, donor behavior, philanthropy, human rights, repression
---

@@

![Causal diagram of the relationship of organizational and structural causes of donor behavior](output/figures/causal-path.pdf){#fig:causal-path}

@@

Issue area contentiousness directly influences legal crackdowns, however, and confounds the heuristic effect of crackdowns on donor behavior (see @fig:causal-path). NGOs working on more contentious human rights issues can be challenging to regimes [@Heiss:2017; @Swiney:2019]. These organizations are often viewed as politically motivated outsiders and run a higher risk of getting expelled [@DPR:2015], increasing the possibility of wasting donor resources. Thus, donors may believe that it is more difficult to improve human rights in the host country and subsequently shy away from donating to NGOs working on political causes.

@@

The final heuristic that can influence individual donors that we examine is the *source of NGO funding*. Government grants to NGOs can often crowd out private funding for nonprofit organizations [@AndreoniPayne:2003; @Brooks:2000], though evidence is mixed on whether this completely or partially crowds out private dollars [@Steinberg:1991; @Heutel:2014]. Donors may feel less inclined to contribute to NGOs receiving funds from agencies like USAID because such funding can signal a lack of independence from the government. Organizations that receive substantial government funding frequently avoid programming that would question the donor government even if it were in line with their mission [@Stroup:2012]. In contrast, individual donors who see that organizations are privately funded may feel that they can also contribute and help.

@@

Importantly, when testing these hypotheses, we do not assume that individual donors have perfect knowledge about an NGO or the countries an NGO works in. Donors do not always know about anti-NGO laws in host countries or who funds an NGO when they make their contributions. Moreover, the average donor will not know exactly what "legal crackdown" entails for any given organization (i.e. does it mean an NGO was expelled, or that its assets were frozen, or that it received a fine?". For the sake of this experiment, this ambiguous definition is not an issue. The heuristics we explore here are explicitly simplified shortcuts to donor information-gathering. Knowing that an organization works with human rights issues (even if they don't know the exact subfield of human rights) should be enough to trigger the issue-based deservingess heuristic. Similarly, knowing that an NGO faces legal trouble should be enough to trigger the structural deservingness heuristic.

@@

We collected two outcome measures: (1) how likely participants would be to donate to IRC (measured with a 5-point scale ranging from "Extremely likely" to "Extremely unlikely"), and (2) how much participants would hypothetically donate to IRC if they had an extra $100. We collapsed the likelihood scale into a binary variable measuring whether the participant is likely ("Extremely likely" and "Somewhat likely") or not likely (all other responses) to donate.[^collapse] We also included an exploratory free response question asking respondents to justify their choice.

@@

```{r child = "output/tables/tbl-likely-diffs.md"}
```

@@

```{r child = "output/tables/tbl-amount-diffs.md"}
```

@@

This research note should be viewed as an exploratory study, and it raises numerous questions for future research. These include determining the extent to which these preferences translate into behavior—here, we have only asked respondents to make a hypothetical donation, but preferences may not reflect actual behavior. Future research can also test the generalizability of our findings to other issue areas and other political contexts. For instance, recent research has found a substantial uptick in anti-NGO legislation in democracies where where private foreign donors reside [@Swiney:2019], and this constricting legal space in donors' own countries will likely influence their giving patterns.

@@

[^supp]: See the online appendix for details about our sample demographics and balance across experimental conditions.

[^collapse]: An ordered probit model in the online appendix shows that results are consistent when using all five possible responses.

[^priors]: See the online appendix for details about our prior choices and computational environment.

[^mturk]: Samples from Mechanical Turk are not nationally representative and our sample skewed younger and more politically liberal. However, because we aimed our study at people who are willing to donate online, we are not overly concerned with these discrepancies. Moreover, a growing body of research comparing Mechanical Turk with other sampling services finds it generally comparable (see the online appendix for a lengthier discussion about this research).

[^targeting]: While anti-NGO legislation can theoretically target all NGOs within a state's borders, in practice these laws are enforced more selectively against contentious groups that pose threats to states [see @Chaudhry:2016; @Heiss:2017].

### 2020-01-21 manuscript/supplement.Rmd
Get rid of extra page
More final-ish edits

MTurk has become an increasingly popular method for fielding survey experiments on cheap semi-representative national samples, though drawing definitive causal inference from convenience samples is not without issues. Previous research has found that MTurk workers tend to be more liberal, younger, and less racially diverse than the general US population [@CliffordJewellWaggoner:2015; @BerniskyHuberLenz:2012]. The results from our experiment generally mirror these findings. @tbl:exp-sample compares demographic characteristics of our sample with national averages from the US Census's Current Population Survey (CPS) [@IPUMS-CPS:2017]. For general demographic information, we use the 2017 Annual Social and Economic Supplement (ASEC) to the CPS. From 2002–2015, the CPS included a Volunteer Supplement every September, so we use 2015 data for data on volunteering and donating to charity. We do not show other respondent demographic details because we do not have good population-level data to compare our sample with. We could theoretically use Pew data to compare political preferences, but Pew collects data on party affiliation, while we collected data about respondents' ideological positions along a conservative–liberal spectrum, which makes the two variables incomparable.

@@

As seen in @tbl:exp-sample, our sample is younger, wealthier, and more educated than national averages. This is to be expected, given previous findings about the characteristics of MTurk workers and given that our target population is the portion of Americans willing to give money to charities online, which implies access to technology and disposable income. Our sample self-reports high levels of charitable giving: 82% said they donate at least once a year, while only 49% donate to charity in the previous year nationally. There are several possible explanations for this discrepancy. First, there are differences in the wording of our question and CPS's question—we ask how often respondents typically donate, while CPS asks if respondents have donated (or not) in the past 12 months. Second, respondents were primed and knew that the survey was related to NGOs, and thus might be displaying social desirability bias. Finally, some charitably-oriented respondents may have self-selected into the survey, given that the title listed on MTurk mentioned international nonprofit organizations. We are not overly concerned with these discrepancies, since we aimed our study at those who would be more inclined to donate online.

@@

## Treatment assignments and CONSORT diagram

@@

@fig:consort uses a CONSORT diagram to show the assignment of participants to the eight different experimental conditions. We excluded participants that (1) did not participate in the experiment through MTurk and (2) failed either attention check.

>>>>

Amazon's Mechanical Turk (MTurk) has become an increasingly popular method for fielding survey experiments on cheap semi-representative national samples, though drawing definitive causal inference from convenience samples is not without issues. Previous research has found that MTurk workers tend to be more liberal, younger, and less racially diverse than the general US population [@CliffordJewellWaggoner:2015; @BerniskyHuberLenz:2012]. The results from our experiment generally mirror these findings, as seen in @tbl:exp-sample, which compares demographic characteristics of our sample with national averages from the US Census's Current Population Survey (CPS) [@IPUMS-CPS:2017]. For general demographic information, we use the 2017 Annual Social and Economic Supplement (ASEC) to the CPS. From 2002–2015, the CPS included a Volunteer Supplement every September, so we use 2015 data for data on volunteering and donating to charity. We do not show other respondent demographic details because we do not have good population-level data to compare our sample with. We could theoretically use Pew data to compare political preferences, but Pew collects data on party affiliation, while we collected data about respondents' ideological positions along a conservative–liberal spectrum, which makes the two variables incomparable.

@@

As seen in @tbl:exp-sample, our sample is younger, wealthier, and more educated than national averages. This is to be expected, given previous findings about the characteristics of MTurk workers and given that our target population is the portion of Americans willing to give money to charities online, which implies access to technology and disposable income. Our sample self-reports high levels of charitable giving: 82% said they donate at least once a year, while only 49% donate to charity in the previous year nationally. There are several possible explanations for this discrepancy. First, there are differences in the wording of our question and CPS's question—we ask how often respondents typically donate, while CPS asks if respondents have donated (or not) in the past 12 months. Second, respondents were primed and knew that the survey was related to NGOs, and thus might be displaying social desirability bias. Finally, some charitably-oriented respondents may have self-selected into the survey, given that the title listed on MTurk mentioned international nonprofit organizations.

We are not overly concerned with these discrepancies, since we aimed our study at those who would be more inclined to donate online. Moreover, in an impressive study to test the validity of experiments conducted on non-representative, internet-based samples, @MullinixLeeperDruckman:2015 ran 20 identical experiments on both MTurk and on a nationally representative population sample. The results from both types of surveys tracked closely with each other and MTurk-based surveys were able to replicate most of the treatment effects found in national samples. Accordingly, they concluded that internet-based population samples are "useful testing grounds for experimental social science" and that they can still lead to substantial progress in generating knowledge and making inferential claims [@MullinixLeeperDruckman:2015, 124]. MTurk-based studies ultimately do not replace the need for more robust nationally representative samples, which still provide the most robust and accurate evidence for causal claims, but they are a cost-effective method for estimating potential population-level effects. Other research corroborates these conclusions—while MTurk samples are not perfectly representative, they are still helpful approximations of representativeness [@BerniskyHuberLenz:2012; @KrupnikovLevine:2014; @CliffordJewellWaggoner:2015; @CoppockLeeperMullinix:2018].

@@

## Treatment assignment balance and CONSORT diagram

@@

@fig:consort and @tbl:experimental-conditions show the assignment of participants to the eight different experimental conditions. We excluded participants that (1) did not participate in the experiment through MTurk and (2) failed either attention check.

```{r child = "output/tables/tbl-experimental-conditions.md"}
```
